Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-09-07 article/article.tex
Commit 2024-09-07 19:12

Some of the proposed underling mechanisms of those enhancements have
been summarized by \citet{Gueymard2017}; the most important being the
scattering of radiation on the edges of cumulus clouds. It has been also
suggested that enhancement of GHI can be produced by thin cirrus clouds
though refraction and scattering \citep{Thuillier2013}. Further
investigation with radiative transfer modeling and observations pointed
as the prevailing mechanism, the strong forward Mie scattering through
clouds of low optical depth
\citep{Pecenak2016, Thuillier2013, Yordanov2013, Yordanov2015}. Overall,
the appearance of CE events depend on different interactive factors,
which include cloud thickness, structure and type, and the relative
position of the sun and the clouds \citep{Gueymard2017, Veerman2022}.

@@

In this study, we evaluate the effects of CE events on GHI by
investigating their frequency of occurrence, intensity, and duration in
a thirty-year record of GHI observations at Thessaloniki, Greece. We
used modeled cloud-free irradiance as a baseline to identify cloud
enhancements, and we determined long-term trends of the above-mentioned
metrics, their climatology and some general characteristics. To our
knowledge there are no other studies that provide trends from such a
long dataset.

@@

To further improve the quality of the irradiance data, a manual
screening was performed, to remove inconsistent and erroneous recordings
that can occur stochastically or systematically during the long
operation of the instrument. The manual screening was aided by a
radiation data quality assurance procedure, adjusted for the site, which
was based on the methods of Long and Shi~\citep{Long2006, Long2008a}.
Thus, problematic recordings have been excluded from further processing.
Furthermore, due to the significant measurement uncertainty in GHI when
the Sun is near the horizon, and due to some systematic obstructions by
nearby buildings, we have excluded all measurements with solar zenith
angle (SZA) greater than \(78^\circ\). Finally, images from a sky camera
have been used in the manual inspection of the CE identification. The
sky camera has been operating since 2012 and stores images in 5 min time
steps. An overview of the GHI data used in this study is given in
Figure\nobreakspace{}\ref{fig:CLB-daily}, as the daily means.

@@

{\centering \includegraphics[width=1\linewidth]{../images/P_daily_trend-3}

@@

In this study, we define an event as CE when the measured GHI at ground
level exceeds the expected value under cloud-free conditions. Similarly,
we define as extreme cloud enhancement events (ECE), the cases when GHI
at ground level exceeds the Total Solar Irradiance (TSI) at the top of
the atmosphere (TOA) for the same SZA. Although the duration of these
bursts can vary from seconds to several minutes, here we are constrained
by the temporal resolution of our data to identify events with duration
of at minimum one-minute.

For the detection of CE cases, we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OI). The OI is defined as the irradiance
difference of the measured one-minute GHI from the
\(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere. First,
we used a simple approach for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple cloud-free model and was already adjusted and applied
to our data \citep{Natsis2023}. We created a threshold by using an
appropriate relative and/or an additional constant offset. The initial
results showed that we can detect a big portion of the actual CE events.
However, by inspecting the daily plots of irradiance it became evident
that changes in atmospheric conditions introduced numerous false
positive or false negative results. The main reason for these
discrepancies is the variability of the effects of aerosols and water
vapor which were not considered in the simple method. To produce a more
representative reference we included the effects of these factors using
a radiative transfer model (RTM). The applied methodology is discussed
in section\nobreakspace{}\ref{rtmcs}.

@@

Main factors responsible for the attenuation of the broadband downward
solar radiation under cloud free atmospheres are aerosols and water
vapor. At Thessaloniki, such measurements are available since 2003 from
a Cimel sun photometer which is part of the Aerosol Robotic Network
(AERONET) \citep{Giles2019, Buis1998}. From the observations in the
period 2003 -- 2023 we calculated the monthly climatological means and
standard deviation (\(\sigma\)) for the aerosol optical depth (AOD) at
\(500\,\text{nm}\) and the equivalent height of the water column (WC).
The monthly climatological values of AOD and WC, as well as combinations
with additional offsets of \(\pm1\sigma\) and \(\pm2\sigma\), were used
as inputs to the RTM in the construction of the LUT.

@@

The cloud-free reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for cloud-free
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using simulations with the RTM to account for the
long-term variations of the AOD. As AERONET data start only in 2003, we
used for the period 1994 -- 2005 estimates of changes in AOD at
\(340\,\text{nm}\) derived from a collocated Brewer spectrophotometer
\citep{Kazadzis2007} to calculate the trend in \(\text{GHI}_\text{ref}\)
due to aerosols during this period.

@@

>>>>

Some of the proposed underling mechanisms of those enhancements have
been summarized by \citet{Gueymard2017}; the most important being the
scattering of radiation on the edges of cumulus clouds. It has also been
suggested that enhancement of GHI can be produced by thin cirrus clouds
though refraction and scattering \citep{Thuillier2013}. Further
investigation with radiative transfer modeling and observations pointed
as the prevailing mechanism, the strong forward Mie scattering through
clouds of low optical depth
\citep{Pecenak2016, Thuillier2013, Yordanov2013, Yordanov2015}. Overall,
the appearance of CE events depend on different interactive factors,
which include cloud thickness, structure and type, and the relative
position of the sun and the clouds \citep{Gueymard2017, Veerman2022}.

@@

In this study, we evaluate the effects of CE events on GHI by
investigating their frequency of occurrence, intensity, and duration in
a thirty-year record of GHI observations at Thessaloniki, Greece. We
used modeled cloud-free irradiance as a baseline to identify cloud
enhancements, and we determined long-term trends of the above-mentioned
metrics, their climatology and some general characteristics. To our
knowledge, there are no other studies that provide trends from such a
long dataset.

@@

To further improve the quality of the irradiance data, a manual
screening was performed, to remove inconsistent and erroneous recordings
that can occur stochastically or systematically during the long
operation of the instrument. The manual screening was aided by a
radiation data quality assurance procedure, adjusted for the site, which
was based on the methods of Long and Shi~\citep{Long2006, Long2008a}.
Thus, problematic recordings have been excluded from further processing.
Furthermore, due to the significant measurement uncertainty in GHI when
the Sun is near the horizon, and due to some systematic obstructions by
nearby buildings, we have excluded all measurements with solar zenith
angle (SZA) greater than \(78^\circ\). Finally, images from a sky camera
have been used in the manual inspection of the CE identification. The
sky camera has been operating since 2012 and stores images in 5 min time
steps. An overview of the GHI data used in this study is given in
Figure\nobreakspace{}\ref{fig:CLB-daily}, as the daily means. Daily
means were calculated only for days with at least \(60\,\%\) of data
availability; however, all available one-minute measurements have been
used for the detection of CE events.

@@

{\centering \includegraphics[width=1\linewidth]{../images/P_daily_trend-5}

@@

In this study, we define an event as CE when the measured GHI at ground
level exceeds the expected value under cloud-free conditions. Similarly,
we define as extreme cloud enhancement events (ECE), the cases when GHI
at ground level exceeds the Total Solar Irradiance (TSI) at the top of
the atmosphere (TOA) for the same SZA. Although the duration of these
bursts can vary from seconds to several minutes, here we are constrained
by the temporal resolution of our data to identify events with a
duration of at minimum one-minute.

For the detection of CE cases, we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OI). The OI is defined as the irradiance
difference of the measured one-minute GHI from the
\(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere. First,
we used a simple approach for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple clear sky radiation model and was already adjusted and
applied to our data \citep{Natsis2023}. We created a threshold by using
an appropriate relative and/or an additional constant offset. The
initial results showed that we can detect a big portion of the actual CE
events. However, by inspecting the daily plots of irradiance it became
evident that changes in atmospheric conditions introduced numerous false
positive or false negative results. The main reason for these
discrepancies is the variability of the effects of aerosols and water
vapor which were not considered in the simple method. To produce a more
representative reference we included the effects of these factors using
a radiative transfer model (RTM). The applied methodology is discussed
in section\nobreakspace{}\ref{rtmcs}.

@@

Main factors responsible for the attenuation of the broadband downward
solar radiation under cloud free atmospheres are aerosols and water
vapor. At Thessaloniki, such measurements are available since 2003 from
a Cimel sun photometer, which is part of the Aerosol Robotic Network
(AERONET) \citep{Giles2019, Buis1998}. From the observations in the
period 2003 -- 2023 we calculated the monthly climatological means and
standard deviation (\(\sigma\)) for the aerosol optical depth (AOD) at
\(500\,\text{nm}\) and the equivalent height of the water column (WC).
The monthly climatological values of AOD and WC, as well as combinations
with additional offsets of \(\pm1\sigma\) and \(\pm2\sigma\), were used
as inputs to the RTM in the construction of the LUT.

@@

The cloud-free reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for cloud-free
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using RTM simulations to account for the long-term
variations of the AOD. High quality AOD data with a Cimel sun-photometer
(part of AERONET) at Thessaloniki start only in 2003, while spectral AOD
measurements using direct-sun observations with a MKIII Brewer
spectrophotometer are available for the period 1997 -- 2017. These data
are taken sporadically during each day and are less dense compared to
the AERONET data. By comparing monthly AOD data at \(340\,\text{nm}\) of
the two instruments for the common periods of operation we adjusted the
Brewer data and filled the missing months of the Cimel record data with
Brewer data. Using the monthly time series of AOD at \(340\,\text{nm}\),
as well as monthly climatological values of the Angstrom exponent and
constant WC of \(15.6\,\text{mm}\) derived from the Cimel record, we
simulated with the RTM the cloud-free GHI at SZA of \(55^\circ\) for
each month in the period 1997 -- 2023. The SZA of \(55^\circ\) was
chosen as representative of all days in the year to get a rough estimate
of the annually averaged change in cloud-free GHI. A second-degree
polynomial fit was applied to the simulated yearly averaged GHI to
derive the long-term change in GHI due to aerosols.

As AERONET data start only in 2003, we used for the period 1994 -- 2005
estimates of changes in AOD at \(340\,\text{nm}\) derived from a
collocated Brewer spectrophotometer \citep{Kazadzis2007} to calculate
the trend in \(\text{GHI}_\text{ref}\) due to aerosols during this
period.

@@

\begin{equation}
\text{AOD change} [\Delta\%] :  -12170 + 12.05 \cdot y + -0.002981 \cdot y^2  \label{eq:AODchange}
\end{equation} where \(y\) the date as a decimal fraction of the year
